---
title: "analysis_profiling"
output: html_document
---
```{r}
library(profvis)
```

```{r}
aggregateProf <- readRDS("dataOutput/profRes/aggregateProf.rds")
joinProf <- readRDS("dataOutput/profRes/joinProf.rds")
```

```{r}
aggregateProf
```

- Most of the execution time is taken by the invariant_cols2 calls. I expected that this function check for each column if it needs to be discarded (if no consensus between values of PSMs that come from the same peptide).

```{r}
joinProf
```

- Most of the execution time is taken by the reduce call, which is expected.


## Code aggregate

```{r}
library(MsCoreUtils)
library(QFeatures)
.aggregateQFeatures <- function(object, fcol, fun, rowDataCols = NULL, ...) {
    ## Copied from the PSMatch package, given that it is not available
    ## on Bioconductor yet.
    .makePeptideProteinVector <- function(m, collapse = ";") {
        stopifnot(inherits(m, "Matrix"))
        vec <- rep(NA_character_, nrow(m))
        for (i in seq_len(nrow(m)))
            vec[i] <- paste(names(which(m[i, ] != 0)), collapse = collapse)
        names(vec) <- rownames(m)
        vec
    }
    if (missing(fcol))
        stop("'fcol' is required.")
    m <- assay(object, 1)
    if(is.null(rowDataCols)) rd <- rowData(object) else rd <- rowData(object)[, rowDataCols]
    if (!fcol %in% names(rd))
        stop("'fcol' not found in the assay's rowData.")
    groupBy <- rd[[fcol]]

    ## Store class of assay i in case it is not a SummarizedExperiment
    ## so that the aggregated assay can be reverted to that class
    .class <- class(object)

    ## Message about NA values is quant/row data
    has_na <- character()
    if (anyNA(m))
        has_na <- c(has_na, "quantitative")
    if (anyNA(rd, recursive = TRUE))
        has_na <- c(has_na, "row")
    if (length(has_na)) {
        msg <- paste(paste("Your", paste(has_na, collapse = " and "),
                           " data contain missing values."),
                     "Please read the relevant section(s) in the",
                     "aggregateFeatures manual page regarding the",
                     "effects of missing values on data aggregation.")
        message(paste(strwrap(msg), collapse = "\n"))
    }

    if (is.vector(groupBy) & !is.list(groupBy)) { ## atomic vectors
        aggregated_assay <- aggregate_by_vector(m, groupBy, fun, ...)
        aggcount_assay <- aggregate_by_vector(m, groupBy, colCounts)
        aggregated_rowdata <- QFeatures::reduceDataFrame(rd, rd[[fcol]],
                                                         simplify = TRUE,
                                                         drop = TRUE,
                                                         count = TRUE)
        assays <- SimpleList(assay = aggregated_assay, aggcounts = aggcount_assay)
        rowdata <- aggregated_rowdata[rownames(aggregated_assay), , drop = FALSE]
    } else if (is(groupBy, "Matrix")) {
        aggregated_assay <- aggregate_by_matrix(m, groupBy, fun, ...)
        ## Remove the adjacency matrix that should be dropped anyway
        rd[[fcol]] <- NULL
        ## Temp variable for unfolding and reducing - removed later
        rd[["._vec_"]] <- .makePeptideProteinVector(groupBy)
        rd <- unfoldDataFrame(rd, "._vec_")
        aggregated_rowdata <- reduceDataFrame(rd, rd[["._vec_"]], drop = TRUE)
        aggregated_rowdata[["._vec_"]] <- NULL
        ## Count the number of peptides per protein
        .n <- apply(groupBy != 0, 2, sum)
        aggregated_rowdata[[".n"]] <- .n[rownames(aggregated_rowdata)]

        assays <- SimpleList(assay = as.matrix(aggregated_assay)) ## to discuss
        rowdata <- aggregated_rowdata[rownames(aggregated_assay), , drop = FALSE]
    } else stop("'fcol' must refer to an atomic vector or a sparse matrix.")
    se <- SummarizedExperiment(assays = assays,
                               colData = colData(object),
                               rowData = rowdata)

    ## If the input objects weren't SummarizedExperiments, then try to
    ## convert the merged assay into that class. If the conversion
    ## fails, keep the SummarizedExperiment, otherwise use the
    ## converted object (see issue #78).
    if (.class != "SummarizedExperiment")
        se <- tryCatch(as(se, .class),
                       error = function(e) se)
    return(se)
}

```

```{r}
aggregateFeatureCopy <- function(object, i, fcol, name = "newAssay",
                   fun = MsCoreUtils::robustSummary, rowDataCols = NULL, ...) {
              if (isEmpty(object))
                  return(object)
              ## Check arguments
              if (any(present <- name %in% names(object)))
                  stop("There's already one or more assays named: '",
                       paste0(name[present], collapse = "', '"), "'.")
              i <- QFeatures:::.normIndex(object, i)
              if (length(i) != length(name)) stop("'i' and 'name' must have same length")
              if (length(fcol) == 1) fcol <- rep(fcol, length(i))
              if (length(i) != length(fcol)) stop("'i' and 'fcol' must have same length")
              ## Aggregate each assay
              for (j in seq_along(i)) {
                  from <- i[[j]]
                  to <- name[[j]]
                  by <- fcol[[j]]
                  ## Create the aggregated assay
                  aggAssay <- .aggregateQFeatures(object[[from]],
                                                  by, fun, rowDataCols, ...)
                  ## Add the assay to the QFeatures object
                  object <- addAssay(object, aggAssay, name = to)
                  ## Link the input assay to the aggregated assay
                  object <- addAssayLink(object, from = from,
                                         to  = to, varFrom = by,
                                         varTo = by)
              }
              object
}



```

```{r}
profvis(aggregateFeatureCopy(PSM1000, i = 1, fcol = "modseq"))


```


```{r}
reduceDataFrame <- function(x, k, count = FALSE,
                            simplify = TRUE, drop = FALSE) {
    res <- split(x, k)
    lens <- unname(lengths(res))
    if (simplify | drop)
        invars <- invariant_cols2(res)
    res <- DataFrame(res)
    if (simplify) {
        ## replace invariant cols
        for (i in invars)
            res[[i]] <- unname(sapply(res[[i]], "[[", 1))
    }
    if (drop)
        res <- res[, invars, drop = FALSE]
    if (count)
        res[[".n"]] <- lens
    res
}
```

```{r}
se <- getWithColData(PSM1000, i = 1)
rd <- rowData(se)
m <- assay(se, 1)
fcol <- "modseq"
profvis({
aggregated_rowdata <- reduceDataFrame(rd, rd[[fcol]],
                                                         simplify = TRUE,
                                                         drop = TRUE,
                                                         count = TRUE)
})
```

```{r}
invariant_col2 <- function(x) all(lengths(lapply(x, unique)) == 1)

invariant_cols2 <- function(x) {
    res <- rep(NA, length(x[[1]]))
    for (i in seq_along(res)) {
        res[i] <- invariant_col2(x[, i])
    }
    which(res)
}

x <- split(rd, rd[[fcol]])
lens <- unname(lengths(res))

profvis(invariant_cols2(res))

t <- res[, 1]
profvis({
    invariant_col2 <- function(x){
        ap <- lapply(x, unique)
        cd <- ap == 1
        ll <- lengths(cd)
        all(ll)
        }

    invariant_col2(t)
    })
```

```{r}
reduceDataFrame2 <- function(x, k, count = FALSE,
                            simplify = TRUE, drop = FALSE) {
    res <- base::split(x, k)
    lens <- unname(lengths(res))
    if (simplify | drop)
        invars <- invariant_cols2(res)
    res <- DataFrame(res)
    if (simplify) {
        ## replace invariant cols
        for (i in invars)
            res[[i]] <- unname(sapply(res[[i]], "[[", 1))
    }
    if (drop)
        res <- res[, invars, drop = FALSE]
    if (count)
        res[[".n"]] <- lens
    res
}
```

```{r}
rd
rd[[fcol]]

tapply(rd[, "Retention.time"], rd[[fcol]], FUN = function(x){length(unique(x))==1})

df <- DataFrame(kept = rep(LETTERS[1:25], 100),
                pep = rep(letters[1:25], 100),
                discard = sample(2500, 2500))

tapply(df, df[["pep"]], FUN = function(x){length(unique(x))==1}, simplify = F)

test1 <- function(df, fcol) {
    res <- rep(NA, ncol(df))
    for (col in seq_along(res)) {
        curr <- all(tapply(df[, col], df[[fcol]], FUN = function(x){length(unique(x))==1}, simplify = T))
        res[col] <- curr
    }
    return(which(res))
}

test2 <- function(df, fcol) {
    which(sapply(colnames(df), function(col) all(tapply(df[, col], df[[fcol]], function(x) length(unique(x)) == 1)), USE.NAMES = F))
}


bmark <- bench::mark(test1(rd, "modseq"),
                     test2(rd, "modseq"),
            {x <- split(rd, rd[[fcol]])
            invariant_cols2(x)})

bmark
```

Could we only test for cols that are still true ?
=> For scp aggregateOverAssays => we can update a vector with the cols marked as TRUE
And only test the others assays for these cols


```{r}
sessionInfo()
```

