---
title: "readSCP Optimisation"
author: "Léopold Guyot"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load data and packages
```{r}
library(QFeatures)
library(scp)
library(microbenchmark)

data("mqScpData")

quantCols <- grep("Reporter.intensity.\\d", colnames(mqScpData),
                   value = TRUE)

data("sampleAnnotation")
```

# Current Implementation

```{r, eval=FALSE}
readSCP <- function(...) {
    ans <- readQFeatures(...)
    el <- ExperimentList(lapply(experiments(ans),
                                as, "SingleCellExperiment"))
    experiments(ans) <- el
    ans
}
```

```{r, message=FALSE}
microbenchmark::microbenchmark(readSCP(assayData = mqScpData,
                                       colData = sampleAnnotation,
                                       runCol = "Raw.file",
                                       removeEmptyCols = TRUE),
                               readQFeatures(assayData = mqScpData,
                                       colData = sampleAnnotation,
                                       runCol = "Raw.file",
                                       removeEmptyCols = TRUE),
                               times = 5)
```

This implementation use readQFeatures to create a QFeatures object that contains
SummarizedExperiment objects, then each SummarizedExperiment is converted into
a SingleCellExperiment. This conversion takes time.

A better approach could be to directly create the QFeatures object with
SingleCellExperiment objects instead of SummarizedExperiment objects.

To achieve that, we need to replace this line in the readSummarizedExperiment
function from QFeatures:

```{r, eval=FALSE}
# old
SummarizedExperiment(assay, rowData = fdata)

#new
SingleCellExperiment(assay, rowData = fdata)
```

```{r, message=FALSE}
# load new readSCP implementation

source(file = "R/readSCP2.R")

microbenchmark::microbenchmark(readSCP(assayData = mqScpData,
                                       colData = sampleAnnotation,
                                       runCol = "Raw.file",
                                       removeEmptyCols = TRUE),
                               readSCP2(assayData = mqScpData,
                                       colData = sampleAnnotation,
                                       runCol = "Raw.file",
                                       removeEmptyCols = TRUE),
                               readQFeatures(assayData = mqScpData,
                                       colData = sampleAnnotation,
                                       runCol = "Raw.file",
                                       removeEmptyCols = TRUE),
                               times = 5)
```

We can see that this new implementation is worse than the current 
implementation. But why?

After multiple profiling of the readSCP2 code, and the different internal 
functions. It appears that processing SingleCellExperiment objects takes way
more time than processing SummarizedExperiment objects. To understand why,
we explore the difference between a simple call to rowData on a sce and a se.

```{r, message=FALSE}
sce <- readRDS("dataOutput/singleCell.rds")
se <- readRDS("dataOutput/summarizedCell.rds")

res <- microbenchmark::microbenchmark(rowData(sce),
                                      rowData(se),
                                      times = 50)
```

We can see that using rowData on a sce takes way more time than on a se.
To investigate why we first look at the source code.

```{r}
showMethods("rowData")

getMethod("rowData", "SingleCellExperiment")

getMethod("rowData", "SummarizedExperiment")
```

We can see that in both case, we call the SummarizedExperiment implementation
(via callNextMethod for sce).
This implementation use the mcols method:

```{r}
showMethods("mcols")

getMethod("mcols", "RangedSummarizedExperiment")

getMethod("mcols", "Vector")
```
Here we can see that a SummarizedExperiment will call the Vector implementation
of the mcols method. While a SingleCellExperiment will call the 
RangedSummarizedExperiment implementation of mcols. This is the case because 
SingleCellExperiment directly inherit from RangedSummarizedExperiment. In fact,
processing sce takes more time than se because it use RangedSummarizedExperiment
methods that use `rowRanges()` which itself call updateObject which takes quite
a lot of computation.
```{r}
microbenchmark(rowRanges(sce), times = 50)
```
With this benchmark we can see that the runtime of rowRanges corresponds to the
difference between calling rowData on a SingleCellExperiment and on a
SummarizedExperiment.


# Conclusion

SingleCellExperiment objects are less efficient than SummarizedExperiment
objects because they inherit from RangedSummarizedExperiment which require
more work for simple tasks (like getting slots'data like the rowData).

The fonctionnalities of SingleCellExperiment are never used in a classical 
workflow, therefore we recommand to use the base readQFeatures function to 
generate the QFeatures object with SummarizedExperiment objects. And when a
SingleCellExperiment object is needed, we can then convert the SummarizedExperiment
object to a SingleCellExperiment object (for instance when using the scplainer
pipeline).

