---
title: "Simulation for PseudoBulk Benchmarking"
author: "Léopold Guyot"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: 
    mathjax: default
---
# Idea

- Subtype effect determined at the patient level?
- Currently a mix of the two, not ideal
- Remove subtype effect
## Packages
```{r packages, message = FALSE}
library(scpdata)
library(scp)
library(SingleCellExperiment)
library(msqrob2)
library(iCOBRA)
library(UpSetR)
library(tidyverse)
```
# Simulation strategy

## Experimental Data
To generate our data, we will use the dataset obtained from Brunner et al. 2022. (Label-Free). We will isolate only one condition (Untreated). For this condition we compute for each protein; his mean log-intensity and his standard deviation

## Intensities modelling

Let \( X_{ij} \) denote the log-intensity of protein \( i \) in cell \( j \).

We model the log-intensity as:

$$
X_{ij} = \mu_i + \epsilon_{ij} + A^{\text{patient}}_i \cdot \delta^{\text{patient}}_{i, p_j} + A^{\text{cell type}}_i \cdot \delta^{\text{cell type}}_{i, c_j}
+ A^{\text{treatment}}_i \cdot \delta^{\text{treatment}}_i \cdot \mathbb{I}_{t_j = 1}
$$

Where:

- \( \mu_i \): baseline log-intensity of protein \( i \) obtained from experimental data
- \( \epsilon_{ij} \sim \mathcal{N}(0, \sigma_i^2) \): residual noise obtained from experimental data
- \( A^{\text{patient}}_i \): indicator if protein \( i \) is affected by patient-specific shifts
- \( A^{\text{patient}}_i \): indicator if protein \( i \) is affected by cell type-specific shifts  
- \( A^{\text{treatment}}_i \): indicator if protein \( i \) is affected by treatment  
- \( p_j \in \{1, \ldots, P\} \): patient identity of cell \( j \)
- \( c_j \in \{1, \ldots, P\} \): cell type identity of cell \( j \)  
- \( t_j \in \{0,1\} \): treatment status of cell \( j \) (1 = treated)  
- \( \delta^{\text{patient}}_{i, p} \sim \mathcal{N}(0, \sigma_{\text{patient}}^2) \): patient-specific shift for protein \( i \) and patient \( p \)
- \( \delta^{\text{cell type}}_{i, c} \sim \mathcal{N}(0, \sigma_{\text{cell type}}^2) \): cell type-specific shift for protein \( i \) and cell type \( p \)  
- \( \delta^{\text{treatment}}_i \sim \mathcal{N}(0, \sigma_{\text{treatment}}^2) \): treatment effect for protein \( i \)  
- \( \mathbb{I}_{t_j = 1} \): indicator function equals 1 if cell \( j \) is treated, 0 otherwise

### Missing Data Modelling

Each \( X_{ij} \) is observed with probability \( 1 - \pi_i \), where:

$$
M_{ij} \sim \text{Bernoulli}(\pi_i)
$$

If \( M_{ij} = 1 \), then \( X_{ij} = \text{NA} \).

Where \( \pi_i \) is the experimentally observed missing rate of protein \( i \) .


## Simulation parameters

```{r parameters}
# number of patients
nPatient = 16

# Cells by patient/subtype
nCellPatientSub = 100


nCellTot = nPatient * nCellPatientSub

# Inter-patient variability
patientSD <- 1

# Prop of proteins shifted by patient effect

patientEffect <- 0.3

nCellTypePerPatient <- 5
nCellPerType <- nCellPatientSub / nCellTypePerPatient

cellTypeEffect <- 0.3       # proportion of proteins affected
cellTypeShiftSD <- 1        # effect size

# proportion of proteins affected by treatment
treatmentEffect <- 0.3

# treatment shift strength
treatmentStrength <- 1  

```

## Experimental Data

Prepare base data obtained from Brunner et al. 2022:
Single cell proteomics data acquired by the Mann Lab using a newly designed timsTOF instrument, referred to as timsTOF-SCP. The dataset contains quantitative information from single-cells blocked at 4 cell cycle stages: G1, G1-S, G2, G2-M. The data was acquired using a label-free sample preparation protocole combined to a data independent (DIA) acquisition mode.

We subset the PSM data for the Untreated cell group.
Then we apply a log2 transformation to have the intensities disitributed normally.
```{r baseData}
base <- scpdata::brunner2022()
sce <- getWithColData(base, "proteins")
sce <- sce[, colData(sce)$CellCycleStage == "UB"]
sce <- logTransform(sce) # log transform to obtain normal intensities distribution
sce <- sce[rowSums(!is.na(assay(sce))) > 1, ]
sce
```

## Intensity distribution
```{r intensityDistReal}
int_mat <- assay(sce)
int_vec <- as.vector(int_mat)
int_vec <- int_vec[!is.na(int_vec)]
ggplot(data.frame(intensity = int_vec), aes(x = intensity)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of protein log-intensities across all cells",
       x = "Intensity", y = "Frequency")
```
## Get log-intensities metrics
```{r intMetrics}
protAssay <- as.data.frame(assay(sce))

# Compute metric
meanIntensity <- rowMeans(protAssay, na.rm = TRUE)
sdIntensity <- apply(protAssay, 1, sd, na.rm = TRUE)
pctMissing <- rowMeans(is.na(protAssay)) * 100

protMetrics <- data.frame(
  protNames = rownames(protAssay),
  meanIntensity = meanIntensity,
  sdIntensity = sdIntensity,
  pctMissing = pctMissing,
  stringsAsFactors = FALSE
)
head(protMetrics)
```
## Intensities metrics distribution

```{r metricsDist}
protMetrics_long <- protMetrics %>%
  pivot_longer(cols = c(meanIntensity, sdIntensity, pctMissing),
               names_to = "metric",
               values_to = "value")

ggplot(protMetrics_long, aes(x = value)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  facet_wrap(~ metric, scales = "free", ncol = 1) +
  theme_minimal() +
  labs(title = "Distribution of per-protein metrics",
       x = "Value", y = "Count")
```

## Simulation (patient and subtype pattern)
```{r extractMetrics}
mu <- protMetrics$meanIntensity
proteinNames <- protMetrics$protNames
residualSD <- protMetrics$sdIntensity
missingProb <- protMetrics$pctMissing/100
nProt <- length(mu)
```

```{r iniMatrix}
# Initialize matrix
simMatrix <- matrix(NA, nrow = nProt, ncol = nCellTot)
rownames(simMatrix) <- proteinNames
cellIds <- paste0("P", rep(1:nPatient, each = nCellPatientSub), "_C", 1:nCellTot)
colnames(simMatrix) <- cellIds
```

```{r shiftProtDef}
nPatientShiftProt <- floor(patientEffect * nProt)
patientShiftProteins <- sample(proteinNames, nPatientShiftProt)

```

```{r prepColData}
# Prepare colData dataframe

patients <- sub("_.*", "", colnames(simMatrix))

colData <- data.frame(
  cellId = colnames(simMatrix),
  patient = patients,
  patientShift = numeric(nCellTot),
  stringsAsFactors = FALSE
)

cellTypes <- character(nCellTot)
for (i in 1:nPatient) {
  for (j in 1:nCellTypePerPatient) {
    startIdx <- ((i - 1) * nCellPatientSub) + ((j - 1) * nCellPerType) + 1
    endIdx <- startIdx + nCellPerType - 1
    cellTypes[startIdx:endIdx] <- paste0("Type", j)
  }
}

colData$cellType <- cellTypes

```

```{r simulation}
cellCounter <- 1

# Simulate
for (i in 1:nPatient) {
  patientShift <- numeric(nProt)
  names(patientShift) <- proteinNames
  patientShift[patientShiftProteins] <- rnorm(nPatientShiftProt, mean = 0, sd = patientSD)
  
  for (k in 1:nCellPatientSub) {
      cellIndex <- cellCounter
      colData$patientShift[cellIndex] <- list(patientShift)
      simMatrix[, cellIndex] <- rnorm(nProt, mean = mu, sd = residualSD) + patientShift 
      
      cellCounter <- cellCounter + 1
    }
  }
```

## Simulate data missingness
```{r simulateNA}
# Add missing values

for (p in seq_len(nProt)) {
  missing <- runif(nCellTot) < missingProb[p]
  simMatrix[p, missing] <- NA
}
```

Should the sampling of shifted proteins be the same for all patients/celltypes?

## Create the SingleCellExperiment object

```{r buildSCE}
simSCE <- SingleCellExperiment(assays = SimpleList(simMatrix),
                     rowData = rowData(sce),
                     colData = colData)
simSCE
```
## Add cell type effect

```{r}
nCellTypeShiftProt <- floor(cellTypeEffect * nProt)
cellTypeShiftProteins <- sample(proteinNames, nCellTypeShiftProt)

rowData(simSCE)$cellTypeAffected <- proteinNames %in% cellTypeShiftProteins
exprMat <- assay(simSCE)

# For each patient
for (i in 1:nPatient) {
  for (j in 1:nCellTypePerPatient) {
    thisCellType <- paste0("Type", j)
    cellsIdx <- which(colData$patient == paste0("P", i) &
                       colData$cellType == thisCellType)
    
    # Define cell-type-specific shift (same for all cells of that type)
    shiftVec <- numeric(nProt)
    names(shiftVec) <- proteinNames
    shiftVec[cellTypeShiftProteins] <- rnorm(nCellTypeShiftProt, mean = 0, sd = cellTypeShiftSD)
    
    # Apply shift to these cells
    exprMat[cellTypeShiftProteins, cellsIdx] <- exprMat[cellTypeShiftProteins, cellsIdx] +
      shiftVec[cellTypeShiftProteins]
  }
}

# Save back to SCE
assay(simSCE) <- exprMat
```

## Add treatment effect

```{r addTreatmentToColData}
# Initialize treatment vector
treatment_vector <- character(ncol(simSCE))
names(treatment_vector) <- colnames(simSCE)

# Get colData
cd <- colData(simSCE)

# Loop over each patient-subtype group
for (p in unique(cd$patient)) {
    # Find cell indices in this group
    group_cells <- which(cd$patient == p)
    
    # Shuffle and assign 50% treated, 50% control
    n_group <- length(group_cells)
    n_treat <- floor(n_group / 2)
    
    treated_cells <- sample(group_cells, size = n_treat)
    control_cells <- setdiff(group_cells, treated_cells)
    
    treatment_vector[treated_cells] <- "treated"
    treatment_vector[control_cells] <- "control"
  }

# Assign to colData
colData(simSCE)$treatment <- treatment_vector
```

```{r addTreatEffect}
# Select affected proteins
nProt <- nrow(simSCE)
proteinNames <- rownames(simSCE)
nAffected <- floor(treatmentEffect * nProt)
affected_proteins <- sample(proteinNames, size = nAffected, replace = FALSE)

# Simulate shift per affected protein
treatment_shift_vec <- numeric(nProt)
names(treatment_shift_vec) <- proteinNames
treatment_shift_vec[affected_proteins] <- rnorm(nAffected, mean = 0, sd = treatmentStrength)

# Modify assay
exprMat <- assay(simSCE)
treated_cells <- colData(simSCE)$treatment == "treated"

# Apply shift: add the same shift for each affected protein across treated cells
exprMat[affected_proteins, treated_cells] <- exprMat[affected_proteins, treated_cells] +
  treatment_shift_vec[affected_proteins]

# Save modified matrix back
assay(simSCE) <- exprMat

rowData(simSCE)$treatmentAffected <- rownames(simSCE) %in% affected_proteins
rowData(simSCE)$treatmentShift <- treatment_shift_vec
```

```{r showData}
simSCE
```

## Plot new distribution of log-intensities

```{r intDistSim}
int_mat <- assay(simSCE)
int_vec <- as.vector(int_mat)
int_vec <- int_vec[!is.na(int_vec)]
ggplot(data.frame(intensity = int_vec), aes(x = intensity)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of protein log-intensities across all cells",
       x = "Intensity", y = "Frequency")
```


## Apply scplainer

```{r scplainer}
scpSIM <- scpModelWorkflow(simSCE, formula = ~ 1 + patient + treatment + cellType, verbose = FALSE)
scpRes <- scpDifferentialAnalysis(
    scpSIM,
    contrasts = list(c("treatment", "control", "treated"))
)[[1]]
pca <- scpComponentAnalysis(scpSIM, method = "APCA", residuals = FALSE, unmodelled = FALSE)
colData(scpSIM)$cell <- colData(scpSIM)$cellId
bySamplePCs <- scpAnnotateResults(
    pca$bySample, colData(scpSIM), by = "cell"
)
scpComponentPlot(bySamplePCs, pointParams = list(
        aes(colour = cellType), 
        alpha = 0.6
    ))
```
## Apply msqrob2
```{r defineContrast}
L <- makeContrast("treatmenttreated=0", parameterNames = c("treatmenttreated"))
```

```{r msqrob2}
msqSIM <- suppressWarnings(msqrob(simSCE, formula = ~ (1 | patient) + treatment + cellType))
msqRes <- rowData(hypothesisTest(object = msqSIM, contrast = L))$treatmenttreated
```

## Apply PseudoBulk + msqrob2

```{r pseudobulk}
aggregate_se <- function(se, group_by_cols, fun = mean) {
    if (!all(group_by_cols %in% colnames(colData(se)))) {
        stop("Some specified group_by_cols do not exist in colData.")
    }
    colData(se)$Group <- apply(colData(se)[, group_by_cols, drop = FALSE], 1, paste, collapse = "_")
    group_factor <- factor(colData(se)$Group)
    assay_matrix <- assay(se)

    # Aggregate assay matrix by group_factor levels order
    grouped_indices <- split(seq_along(group_factor), group_factor)
    levels_group <- levels(group_factor)  # factor levels in order
    aggregated_assay <- do.call(cbind, lapply(levels_group,
                                              function(g) {
                                                  idx <- grouped_indices[[g]]
                                                  apply(assay_matrix[, idx, drop = FALSE], 1, fun, na.rm = TRUE)
                                              }))
    new_colData <- colData(se) %>%
        as.data.frame() %>%
        group_by(Group) %>%
        summarise(across(all_of(group_by_cols), dplyr::first), .groups = "drop") %>%
        as.data.frame()
    # Reorder new_colData rows to match factor levels
    new_colData <- new_colData[match(levels_group, new_colData$Group), ]
    rownames(new_colData) <- new_colData$Group
    new_se <- SummarizedExperiment(assays = list(aggregated_assay),
                                   colData = new_colData,
                                   rowData = rowData(se))
    return(new_se)
}


pbSIM <- aggregate_se(simSCE, c("patient", "treatment", "cellType"), fun = mean)
pbSIM <- suppressWarnings(msqrob(pbSIM, ~ 1 + patient + treatment + cellType))
pbRes <- rowData(hypothesisTest(object = pbSIM, contrast = L))$treatmenttreated
```

# Get FDR curves

```{r pvalDF}
# Extract p-values from each result
pval_pb <- pbRes[, "pval", drop = FALSE]
colnames(pval_pb) <- "PBmsqrob2"

pval_msq <- msqRes[, "pval", drop = FALSE]
colnames(pval_msq) <- "msqrob2"

pval_scp <- as.data.frame(scpRes[, c("feature", "pvalue")])
colnames(pval_scp) <- c("feature", "scplainer")
rownames(pval_scp) <- pval_scp$feature
pval_scp$feature <- NULL

# Merge all three by rownames (feature names)
tmp1 <- merge(pval_pb, pval_msq, by = "row.names", all = TRUE)
rownames(tmp1) <- tmp1$Row.names
tmp1$Row.names <- NULL

combined_pvals <- merge(tmp1, pval_scp, by = "row.names", all = TRUE)
rownames(combined_pvals) <- combined_pvals$Row.names
combined_pvals$Row.names <- NULL

# Extract adjusted p-values from each result
padj_pb <- pbRes[, "adjPval", drop = FALSE]
colnames(padj_pb) <- "PBmsqrob2"

padj_msq <- msqRes[, "adjPval", drop = FALSE]
colnames(padj_msq) <- "msqrob2"

padj_scp <- as.data.frame(scpRes[, c("feature", "padj")])
colnames(padj_scp) <- c("feature", "scplainer")
rownames(padj_scp) <- padj_scp$feature
padj_scp$feature <- NULL

# Merge all three by rownames (feature names)
tmp_padj <- merge(padj_pb, padj_msq, by = "row.names", all = TRUE)
rownames(tmp_padj) <- tmp_padj$Row.names
tmp_padj$Row.names <- NULL

combined_padj <- merge(tmp_padj, padj_scp, by = "row.names", all = TRUE)
rownames(combined_padj) <- combined_padj$Row.names
combined_padj$Row.names <- NULL
```
```{r truth}
truth <- data.frame(isTreated = rowData(simSCE)$treatmentAffected,
                    row.names = rownames(rowData(simSCE))
                    )
```

```{r COBRAData}
cobData <- COBRAData(
  pval = combined_pvals,
  padj = combined_padj,
  truth = truth
)

perfRes <- calculate_performance(cobradata = cobData,
                      binary_truth = "isTreated") %>%
    fdrtpr() %>%
    mutate(thr = as.numeric(sub("thr", "", thr)))
```
```{r plotFDRTPR}
plot <- ggplot(perfRes, aes(x = FDR, y = TPR, color = method)) +
    geom_vline(
        xintercept = c(0.01, 0.05, 0.1),
        linetype = "dashed", color = "grey50", size = 0.3
    ) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_line(size = 0.7) +
    scale_x_continuous(
        trans = "sqrt",
        limits = c(0, 1),
        breaks = c(0.01, 0.05, 0.2, 0.4, 0.8, 1),
        labels = scales::label_number()
    ) +
    scale_y_continuous(
        limits = c(0, 1),
        breaks = seq(0, 1, 0.2)
    ) +
    theme_minimal() +
    labs(
        title = "FDR vs TPR (square root x-axis)",
        x = "False Discovery Rate (sqrt scale)",
        y = "True Positive Rate",
        color = "Method"
    )
plot
```

## Upset plot

```{r upsetPlot}
scpFeatures <- scpRes %>%
    as.data.frame() %>%
    filter(padj <= 0.05) %>% 
    pull(feature)

msqFeatures <- msqRes %>%
    as.data.frame() %>%
    filter(adjPval <= 0.05) %>% 
    rownames()

pbFeatures <- pbRes %>%
    as.data.frame() %>%
    filter(adjPval <= 0.05) %>% 
    rownames()

truthFeatures <- rowData(simSCE) %>%
    as.data.frame() %>%
    filter(treatmentAffected) %>% 
    rownames()

listUpSet <- list("scplainer" = scpFeatures,
                  "msqrob2" = msqFeatures,
                  "PseudoBulk" = pbFeatures,
                  "truth" = truthFeatures)
upset(fromList(listUpSet), order.by = "freq")
```


```{r session}
sessionInfo()
```

