
## Packages
```{r}
library(scpdata)
library(scp)
library(tidyverse)
```
## Simulation parameters

```{r}
set.seed(123)
# number of patients
nPatient = 4
# Cells by patient
nCellPatient = 50

nCellTot = nPatient * nCellPatient
# Inter-patient variability
patientSD <- 0.3
```

## Data

Prepare base data obtained from Brunner et al. 2022:
Single cell proteomics data acquired by the Mann Lab using a newly designed timsTOF instrument, referred to as timsTOF-SCP. The dataset contains quantitative information from single-cells blocked at 4 cell cycle stages: G1, G1-S, G2, G2-M. The data was acquired using a label-free sample preparation protocole combined to a data independent (DIA) acquisition mode.

We subset the PSM data for the Untreated cell group.
Then we apply a log2 transformation to have the intensities disitributed normally.
```{r}
base <- scpdata::brunner2022()
sce <- getWithColData(base, "proteins")
sce <- sce[, colData(sce)$CellCycleStage == "UB"]
sce <- logTransform(sce) # log transform to obtain normal intensities distribution

```

## Intensity distribution
```{r}
int_mat <- assay(sce)

# 2. Convert to a vector (flatten all protein × cell values)
int_vec <- as.vector(int_mat)

# 3. Remove missing values
int_vec <- int_vec[!is.na(int_vec)]

# 4. Plot distribution
ggplot(data.frame(intensity = int_vec), aes(x = intensity)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of protein intensities across all cells",
       x = "Intensity", y = "Frequency")
```
## Get intensities metrics
```{r}
protAssay <- as.data.frame(assay(sce))

# Compute metric
meanIntensity <- rowMeans(protAssay, na.rm = TRUE)
sdIntensity <- apply(protAssay, 1, sd, na.rm = TRUE)
pctMissing <- rowMeans(is.na(protAssay)) * 100

# Add metrics and feature names into one data frame
protMetrics <- data.frame(
  protNames = rownames(protAssay),
  meanIntensity = meanIntensity,
  sdIntensity = sdIntensity,
  pctMissing = pctMissing,
  stringsAsFactors = FALSE
)
head(protMetrics)
```
## Intensities metrics distribution

```{r}
protMetrics_long <- protMetrics %>%
  pivot_longer(cols = c(meanIntensity, sdIntensity, pctMissing),
               names_to = "metric",
               values_to = "value")

# Plot distribution for each metric
ggplot(protMetrics_long, aes(x = value)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  facet_wrap(~ metric, scales = "free", ncol = 1) +
  theme_minimal() +
  labs(title = "Distribution of per-protein metrics",
       x = "Value", y = "Count")
```
## Simulation

```{r}
# Extract mean values
mu <- protMetrics$meanIntensity
proteinNames <- protMetrics$protNames
residualSD <- protMetrics$sdIntensity
missingProb <- protMetrics$pctMissing/100
nProt <- length(mu)

# Initialize matrix
simMatrix <- matrix(NA, nrow = nProt, ncol = nCellTot)
rownames(simMatrix) <- proteinNames
cellIds <- paste0("P", rep(1:nPatient, each = nCellPatient), "_C", 1:nCellTot)
colnames(simMatrix) <- cellIds

# Prepare colData dataframe

colData <- data.frame(
  cellId = colnames(simMatrix),
  patient = sub("_.*", "", colnames(simMatrix)),
  patientShift = numeric(nCellTot),
  stringsAsFactors = FALSE
)

cellCounter <- 1
# Simulate
for (i in 1:nPatient) {
  # simulate patient effect
  patientShift <- rnorm(nProt, mean = 0, sd = patientSD)
  
  for (j in 1:nCellPatient) {
    cellIndex <- (i - 1) * nCellPatient + j
    colData$patientShift[cellCounter] <- patientShift
    # simulate noise
    noise <- rnorm(nProt, mean = 0, sd = residualSD)
    
    # final simulated intensity
    simMatrix[, cellIndex] <- mu + patientShift + noise
    
    cellCounter <- cellCounter + 1
  }
}

# Add missing values

for (p in seq_len(nProt)) {
  missing <- runif(nCellTot) < missingProb[p]
  simMatrix[p, missing] <- NA
}
```

