---
title: "Results of the 4 variables benchmark"
author: "Léopold Guyot"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(plotly)
library(DT)
```

# Multi dimension size benchmark

Size report on generated datasets. Each generated dataset is a combination of 4 variables that impact the object size, the variable are the following:
- nFeat = Number of features by cell (PSM identification depth)
- nCell = Number of cells by assay (batch) => 1, 16 or 32
- nAssay = Number of assay (batch) in the QFeatures object
- nCol = Number of columns in the rowData DataFrame (data associated with each PSM)

The goal is to understand what is the most impactful variable in term of object size. We also want to know if the number of cells by assay (ie. Label-Free vs Labelled) impact the size for a constant number of total cells.

```{r loadData}
convert_bytes_to_mib <- function(bytes) {
  mib <- bytes / (2^20)
  return(mib)
}

sizeTable <- read.table("dataOutput/4_variables_benchmark/qfeatures_size_report.tsv", header = TRUE)

sizeCols <- grep("size", colnames(sizeTable))
for (sizeCol in sizeCols) {
    sizeTable[[paste0(colnames(sizeTable)[[sizeCol]], "_MiB")]] <- convert_bytes_to_mib(sizeTable[[sizeCol]])
}
sizeTable$totalCells <- sizeTable$nCell * sizeTable$nAssay
DT::datatable(sizeTable,
                extensions = "FixedColumns",
                options = list(
                    searching = FALSE,
                    scrollX = TRUE,
                    fixedColumns = TRUE,
                    pageLength = 10,
                    lengthMenu = c(10, 25, 50)
                ))
```

